---
title: "Linear Regression R"
author: "Sai Prakash.Donka"
date: "2025-02-20"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)

install.packages("ISLR2")
library(ISLR2)
```

# Boston Dataset Analysis

### Objective

How can we predict the Median Value of Owner_Occupied Homes using the lower

\<what are we analysing? Why? What insight can we gain from this analysis.

### Data Understand and preparation

\<What kinds of variables do we have? What kind of questions can we answer with this data?\>

### Data Loading

\<What does the summary say about this date?\>

```{r load.data}
data(Boston)
```

### Data exploration

```{r missing values}
missing_values <- Boston %>%
  summarise(across(everything(), ~ sum(is.na(.)), .names = "missing_{.col}"))
print(missing_values)
```

### Train-Test Split

\<How does this technique aid our analysis, especially given new data?\>

```{r}
set.seed(123)  

Boston <- Boston %>%
  mutate(id = row_number())

train_data <- Boston %>%
  sample_frac(0.75)

test_data <- anti_join(Boston, train_data, by = "id")


train_data <- select(train_data, -id)
test_data <- select(test_data, -id)
```

### Exploratory Data Analysis

\<what figures did we build?Why? What information do they convey? How it is important to the analysis?\>

```{r histogram for medv}
ggplot(Boston, aes(x = medv)) +
  geom_histogram(fill = "red", binwidth = 2, color = "blue") +
  labs(title = "Distribution of median home values",
       x = "Median ($1000s)",
       y = "Count")

```

```{r LSTAT vs MEDV Scatterplot}
ggplot(Boston, aes(x = lstat, y=medv)) +
  geom_point(alpha = 0.7, color = 'yellow') +
  labs(title = "Scatterplot: LSTAT vs. MEDV",
       x = "Lower Status Population",
       y = "Median home value ($1000s)")

```

### Model Implementation & Explanation

\<what model are we using? why does this/these model(s) apply to the data?What are the pros & cons of this type of model? \>

### Perform simple Linear Regression on Training Data

\<Describe the function & model fit. maybe talk about the evaluation metrics?

\>

```{r Liunear Regression}
lm.fit = lm(medv ~ lstat, data = train_data)
summary(lm.fit)
```

Could built a scatter plot with this regression line onto it.

### Apply Model to Text Data

\<could interpret the Test MSE\>

```{r apply model to test_data}
train_mse <- mean((train_data$medv - predict(lm.fit, newdata = train_data))^2)
test_mse <- mean((test_data$medv - predict(lm.fit, newdata = test_data))^2)

print(paste("Training MSE:", round(train_mse, 2)))
print(paste("Test MSE:", round(test_mse, 2)))
```

### Simple Linear Regression results & interpretation

\<Overall, how good is this fit? What does it say about the data and the question being asked?\>

### Perform Multiple Linear Regression on Training Data

\<what question does this model answer?\>

```{r}
lm.multiple.fit = lm(medv ~ lstat + age, data = train_data)  
summary(lm.multiple.fit)
```

## NHANES Data Analysis

## Objective

please predict BMI using Age, SmokeNow, PhysActive for induviduals between the ages of 18 and 70

## Data understanding & Preparation

## Data loading

```{r}
#install.packages("NHANES")
library(NHANES)
data(NHANES)

SMOKERS = NHANES %>%
  select(BMI, Age, SmokeNow, PhysActive) %>%
  filter(Age >= 18 & Age <= 70)
```


## Data Exploration

```{r}

missing_values <- SMOKERS %>%
  summarise(across(everything(), ~ sum(is.na(.)), .names = "missing_{.col}"))

print(missing_values)
```

## Train - Test Split

```{r}

set.seed(123)


SMOKERS <- SMOKERS %>% mutate(id = row_number())

train_data <- SMOKERS %>% sample_frac(0.75)  
test_data <- anti_join(SMOKERS, train_data, by = "id")


cat("Training Data Size:", nrow(train_data), "\n")
cat("Testing Data Size:", nrow(test_data), "\n")
train_data <- SMOKERS_split
```

## Exploratory Data Analysis

```{r}

library(ggplot2)

ggplot(SMOKERS, aes(x = BMI)) +
  geom_histogram(fill = "white", binwidth = 5, color = "steelblue") +
  labs(title = "Distribution of BMI", x = "bmi", y = "count")
```

## Scatterplot: Age vs. BMI

This helps in visualizing the relationship between **Age** and **BMI**.

```{r}
ggplot(SMOKERS, aes(x = Age, y = BMI)) +
  geom_point(alpha = 0.7, color = 'red') +
  labs(title = "Scatterplot: Age vs. BMI", x = "Age", y = "BMI")
```

## Scatterplot: SmokeNow vs. BMI

```{r}

ggplot(SMOKERS, aes(x = SmokeNow, y = BMI)) +
  geom_boxplot(fill = "steelblue", color = "blue") +
  labs(title = "Boxplot: SmokeNow vs. BMI", x = "SmokeNow", y = "BMI")
```

## Model Implementation

### Perform simple Linear Regression on Training Data

```{r}

lm.fit <- lm(BMI ~ Age, data = train_data)


summary(lm.fit)
```

### Apply Model to Text Data

```{r}
train_mse <- mean((train_data$BMI - predict(lm.fit, train_data))^2)
test_mse <- mean((test_data$BMI - predict(lm.fit, test_data))^2)

print(paste("Training MSE:", round(train_mse, 2)))
print(paste("Test MSE:", round(test_mse, 2)))
```

### Simple Linear Regression results & interpretation

### Perform Multiple Linear Regression on Training Data

```{r}
lm.multiple.fit = lm(BMI ~ Age + SmokeNow + PhysActive, data = train_data)  
summary(lm.multiple.fit)
```

### Apply the Model to Test Data

```{r}
train_mse <- mean((train_data$BMI - predict(lm.multiple.fit, train_data))^2)
test_mse <- mean((test_data$BMI - predict(lm.multiple.fit, test_data))^2)


print(paste("Training MSE:", round(train_mse, 2)))
print(paste("Test MSE:", round(test_mse, 2)))

```
